Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-03-21 Paper/manuscript_v2/manuscript_07_01.tex
updated manuscript_v2, online material and added case studies codes

We have implemented the approach in the probabilistic programming language Stan \citep{carpenter2017stan} as well as subsequently in the \textit{brms} package \citep{burkner2017brms} of the R software \citep{R2019R}. Several illustrative examples of the performance and applicability of the method are shown using both simulated and real datasets. All examples are accompanied by the corresponding Stan and R code.

@@

The remainder of the paper is structured as follows. In Section \ref{ch4_gp}, we introduce GPs with the covariance and spectral density functions. In Section \ref{ch5_sec_method}, the reduced rank approximation to GPs proposed by \cite{solin2018hilbert} is described. In Section \ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit
exact and approximate GPs to real and simulated data are provided in Sections \ref{ch5_sec_univariate_cases} and \ref{ch5_sec_multivariate_cases}. A brief conclusion of the work is made in Section \ref{ch5_sec_conclusion}. Appendix \ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind this Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. 
Online supplemental material with more case studies illustrating the performance and applicability of the method can be found at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \texttt{Paper/online\_appendices}.

@@

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\nu}K_{\nu}\left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}

\noindent where $\nu$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp\bigg(-\frac{1}{2} \frac{r^2}{\ell^2}\bigg),  \\
k_{\frac{3}{2}}(r)&=\alpha\bigg(1+\frac{\sqrt{3}r}{\ell}\bigg) \exp\bigg(-\frac{\sqrt{3}r}{\ell}\bigg). 
\end{align*}

\noindent The former is commonly known as squared exponential (exponentiated quadratic) covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form:
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \exp\bigg(-\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\bigg),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \Bigg(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\Bigg)\exp\Bigg(- \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\Bigg). \\
\end{align*}

@@

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions is given by
%
\begin{align*}
S_{\nu}(\omega)&= \alpha \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\omega^2 \right)^{(\nu+D/2)}
\end{align*}

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ is a distance in the frequency domain, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form:
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D  \exp\left(-0.5 \ell^2 \omega^2\right), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\omega^2 \right)^{-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}

\noindent Particularizing to an input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form:
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \mathrm{exp}\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \right),   \\
S_{\frac{3}{2}}(\omega)&= \alpha 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \right)^{-3}.
\end{align*}

@@

\noindent The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalues problem in equation (\ref{eq_eigenproblem}) are sinusoidal functions. Independently of the covariance function, they can be computed as
%
\begin{align}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^2, \label{eq_eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(x+L)\right). \label{eq_eigenfunction}
\end{align}

If we truncate the sum in (\ref{eq_approxcov}) to the first $m$ terms, the approximate covariance function becomes
%
\begin{equation}
k(x,x') \approx \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x') = \bm{\phi}(x)^\intercal \Delta \bm{\phi}(x'), \nonumber
\end{equation}

\noindent where $\bm{\phi}(x)=\{\phi_j(x)\}_{j=1}^{m} \in {\rm I\!R}^{m}$ is the column vector of basis functions, and $\Delta  \in {\rm I\!R}^{m\times m}$ is the diagonal matrix of the spectral densities $S_{\theta}(\sqrt{\lambda_j})$: 
%
\begin{eqnarray}
\Delta &=&  \begin{bmatrix}
    S_{\theta}(\sqrt{\lambda_1}) & & \\
    & \ddots & \nonumber \\
    & & S_{\theta}(\sqrt{\lambda_m}) \\
  \end{bmatrix}.
\end{eqnarray}

@@

\noindent where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$:
%
\begin{eqnarray}
\Phi &=&  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{eqnarray}

@@

\noindent This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j}^m \left( S_{\theta}(\sqrt{\lambda_j})\right)^{1/2} \phi_j(x) \beta_j,
\end{equation}

@@

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\bm{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\bm{\cdotp}}$. Thus, for $\bm{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,\ldots,m^{\ast}$, we have: 
%
\begin{align}
\bm{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \right\}_{d=1}^D, \label{eq_eigenvalue_multi} \\
%
\phi^{\ast}_j(\bm{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right). \label{eq_eigenfunction_multi}
\end{align}

\noindent The approximate covariance function is then represented as
%
\begin{equation}\label{eq_approxcov_multi}
k(\bm{x},\bm{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\bm{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\bm{x}) \phi^{\ast}_j(\bm{x}'),
\end{equation}

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section \ref{ch4_sec_cov}). We can now write the approximate series expansion of the multivariate function $f$ as
%
\begin{equation}\label{eq_approxf_multi}
f(\bm{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\bm{\lambda}^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\bm{x}) \beta_j, 
\end{equation}

@@

As explained in Section \ref{ch5_sec_method}, the approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, for instance in a one-dimensional input space $\Omega=[-L,L]\subset {\rm I\!R}$:
%
\begin{equation*}
k(\tau) = \sum_{j=1}^{\infty} S_{\theta} \left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0), 
\end{equation*}

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}) for the unidimensional case and in equations (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{eqnarray}\label{eq_diff_covs}
\int k(\tau) \, d\tau - 
\int \sum_{j=1}^m S_{\theta}\left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0) \, d\tau < \varepsilon.
\end{eqnarray}

@@

Additionally, there is a clear relation of the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} depicts how these three factors interact with each other in relation to a close approximation of the HSGP model, in the cases of a GP with square exponential covariance function and Mat\'ern ($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a square exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum $m$ required to achieve a close approximation in terms of satisfying equation (\ref{eq_diff_covs}). Similarly for Figure \ref{fig6_lscale_vs_J_vs_c_Matern} in the case of a Mat\'ern ($\nu$=3/2) covariance function. We have considered an approximation to be a close enough when the difference between densities of the approximate covariance function and the exact covariance function, $\varepsilon$ in equation (\ref{eq_diff_covs}), is below 1$\%$ of the density of the exact covariance function:
%
\begin{eqnarray*}
 \frac{\varepsilon}{\int k(\tau) \, d\tau} < 0.01.
\end{eqnarray*}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_zoomin.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ ($c = \frac{L}{S}$) and the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a square exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_Matern.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_zoomin_Matern.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ ($c = \frac{L}{S}$) and the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a Mat\`ern($\nu$=3/2) covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c_Matern}
\end{figure}

@@

%\section{Univariate case studies}\label{ch5_sec_univariate_cases}
\section{Case studies}\label{ch5_sec_cases}

@@

The regular GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{eqnarray*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \mathcal{N}(0, \sigma^2  I) \\
f(x) &\sim \mathcal{GP}(0, k(x, x', \theta)),
\end{split}
\end{eqnarray*}

@@

Figure \ref{ch5_fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The HSGP model is on average roughly 400 times faster than the regular GP, in this particular application.

The Stan model code for the exact GP, the approximate GP and the spline models of this case study can be found online at \url{https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-I_1D-Simulated-data}.
%

@@

@@

\noindent The mean function $\mu(t)$ will be defined as an additive model in the form: 
%
\begin{eqnarray} \label{ch5_eq_mean_brithday}
\mu(t) = f_1(t) + f_2(t) + f_3(t) + f_4(t).
\end{eqnarray}

The component $f_1(t)$ represents the long-term trends modeled by a GP with squared exponential covariance function,
%
\begin{eqnarray*}
f_1(t) \sim \mathcal{GP}(0,k_1), \hspace{5mm} k_1(t,t') = \sigma_1^2 \exp\bigg(-\frac{1}{2} \frac{(t-t')^2}{\ell_1^2}\bigg), 
\end{eqnarray*}

@@

The component $f_4(t)$ represents the special days effects, modeled as a horse-shoe prior model \citep{piironen2017sparsity}:
%
\begin{eqnarray*}
f_4(t) \sim \mathcal{N}(0,\lambda^2_t \tau^2), \hspace{5mm} \lambda^2_t \sim \mathcal{C}^{+}(0,1).
\end{eqnarray*}

>>>>

We have implemented the approach in the probabilistic programming language Stan \citep{carpenter2017stan} as well as subsequently in the \textit{brms} package \citep{burkner2017brms} of the R software \citep{R2019R}. Several illustrative examples of the performance and applicability of the method are shown using both simulated and real datasets. All examples are accompanied by the corresponding Stan %and R 
code.

@@

The remainder of the paper is structured as follows. In Section \ref{ch4_gp}, we introduce GPs with the covariance and spectral density functions. In Section \ref{ch5_sec_method}, the reduced rank approximation to GPs proposed by \cite{solin2018hilbert} is described. In Section \ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit
exact and approximate GPs to real and simulated data are provided in Section \ref{ch5_sec_cases}. A brief conclusion of the work is made in Section \ref{ch5_sec_conclusion}. Appendix \ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind this Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. 
Online supplemental material with more case studies illustrating the performance and applicability of the method can be found at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.

@@

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\Big(\frac{\sqrt{2\nu}r}{\ell}\Big)^{\!\nu} \! K_{\nu} \Big(\frac{\sqrt{2\nu}r}{\ell}\Big),
\end{align*}

\noindent where $\nu$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp\!\! \big(\! -\frac{1}{2} \frac{r^2}{\ell^2}\big),  \\
k_{\frac{3}{2}}(r)&=\alpha\Big(1+\frac{\sqrt{3}r}{\ell}\Big) \! \exp\!\! \big(\! -\frac{\sqrt{3}r}{\ell}\big). 
\end{align*}

\noindent The former is commonly known as squared exponential (exponentiated quadratic) covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form:
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \exp\!\!\Big(\! -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\Big),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \bigg(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\bigg) \! \exp\!\!\Big(\! - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\Big). \\
\end{align*}

@@

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions is given by
%
\begin{align*}
S_{\nu}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, l^{2\nu}}\Big(\frac{2\nu}{l^2}+4\pi^2\omega^2 \Big)^{\!(\nu+D/2)}
\end{align*}

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ is a distance in the frequency domain, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form:
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D \exp(-0.5 \ell^2 \omega^2), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\,\ell^3}\Big(\frac{3}{\ell^2}+\omega^2 \Big)^{\!-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}

\noindent Particularizing to an input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form:
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \exp\!\!\big(\!-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \big),   \\
S_{\frac{3}{2}}(\omega)&= \alpha \, 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\big(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \big)^{\!-3}.
\end{align*}

@@

\noindent The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalues problem in equation (\ref{eq_eigenproblem}) are sinusoidal functions. Independently of the covariance function, they can be computed as
%
\begin{align}
\lambda_j&=\Big(\frac{j\pi}{2L}\Big)^{\!2}, \label{eq_eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}}\, \sin\!\!\big(\sqrt{\lambda_j}(x+L)\big). \label{eq_eigenfunction}
\end{align}

If we truncate the sum in (\ref{eq_approxcov}) to the first $m$ terms, the approximate covariance function becomes
%
\begin{equation}
k(x,x') \approx \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x') = \bm{\phi}(x)^\intercal \Delta \, \bm{\phi}(x'), \nonumber
\end{equation}

\noindent where $\bm{\phi}(x)=\{\phi_j(x)\}_{j=1}^{m} \in {\rm I\!R}^{m}$ is the column vector of basis functions, and $\Delta  \in {\rm I\!R}^{m\times m}$ is the diagonal matrix of the spectral densities $S_{\theta}(\sqrt{\lambda_j})$: 
%
\begin{align}
\Delta =  \begin{bmatrix}
    S_{\theta}(\sqrt{\lambda_1}) & & \\
    & \ddots & \nonumber \\
    & & S_{\theta}(\sqrt{\lambda_m}) \\
  \end{bmatrix}.
\end{align}

@@

\noindent where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$:
%
\begin{align}
\Phi =  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{align}

@@

\noindent This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j}^m \big( S_{\theta}(\sqrt{\lambda_j})\big)^{\! \frac{1}{2}} \phi_j(x) \beta_j,
\end{equation}

@@

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\bm{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\bm{\cdotp}}$. Thus, for $\bm{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,\ldots,m^{\ast}$, we have: 
%
\begin{align}
\bm{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \Big\{ \Big(\frac{\pi \mathbb{S}_{jd}}{2L_d}\Big)^{\!2} \Big\}_{d=1}^D, \label{eq_eigenvalue_multi} \\
%
\phi^{\ast}_j(\bm{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \sin\!\!\big(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\big). \label{eq_eigenfunction_multi}
\end{align}

\noindent The approximate covariance function is then represented as
%
\begin{equation}\label{eq_approxcov_multi}
k(\bm{x},\bm{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\big(\sqrt{\bm{\lambda}^{\ast}_j}\big)
\phi^{\ast}_j(\bm{x}) \phi^{\ast}_j(\bm{x}'),
\end{equation}

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section \ref{ch4_sec_cov}). We can now write the approximate series expansion of the multivariate function $f$ as
%
\begin{equation}\label{eq_approxf_multi}
f(\bm{x}) \approx \sum_{j=1}^{m^{\ast}} 
\big( S^{\ast}_{\theta} \big(\sqrt{\bm{\lambda}^{\ast}_j} \big)\big)^{\! \frac{1}{2}} \phi^{\ast}_j(\bm{x}) \beta_j, 
\end{equation}

@@

As explained in Section \ref{ch5_sec_method}, the approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, for instance in a one-dimensional input space $\Omega=[-L,L]\subset {\rm I\!R}$:
%
\begin{equation*}
k(\tau) = \sum_{j=1}^{\infty} S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0), 
\end{equation*}

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}) for the unidimensional case and in equations (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{equation}\label{eq_diff_covs}
\Big( \int k(\tau) \mathrm{d}\tau - 
\int \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0) \mathrm{d}\tau \Big) < \varepsilon.
\end{equation}

@@

Additionally, there is a clear relation of the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} depicts how these three factors interact with each other in relation to a close approximation of the HSGP model, in the cases of a GP with square exponential covariance function and Mat\'ern($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a square exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum $m$ required to achieve a close approximation in terms of satisfying equation (\ref{eq_diff_covs}). Similarly for Figure \ref{fig6_lscale_vs_J_vs_c_Matern} in the case of a Mat\'ern($\nu$=3/2) covariance function. We have considered an approximation to be a close enough when the difference between densities of the approximate covariance function and the exact covariance function, $\varepsilon$ in equation (\ref{eq_diff_covs}), is below 1$\%$ of the density of the exact covariance function:
%
\begin{equation*}
 \frac{\varepsilon}{\int k(\tau) \mathrm{d}\tau} < 0.01.
\end{equation*}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_zoomin.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a square exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_Matern.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_zoomin_Matern.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a Mat\`ern($\nu$=3/2) covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c_Matern}
\end{figure}

@@

\section{Case studies}\label{ch5_sec_cases}

@@

The regular GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \mathcal{N}(0, \sigma^2  I) \\
f(x) &\sim \mathcal{GP}(0, k(x, x', \theta)),
\end{split}
\end{align*}

@@

Figure \ref{ch5_fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. {\color{blue}The HSGP model is on average roughly 400 times faster than the regular GP and 10 times faster than the spline model, for this particular application with a univariate input space. The increase of computation as function of the number of basis functions in a univariate input space is relatively slight, as can be seen in the figure.}

The Stan model code for the exact GP, the approximate GP and the spline models of this case study can be found online at {\small \url{ https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_1D-Simulated-data}}.
%

@@

{\color{blue}In the online supplemental material, some other case studies are presented. From those examples, it can be seen how computation demand of the HSGP model increases severely with the number of input dimensions ($D$) since the number of basis functions in the approximation increases exponentially with $D$ (see equation (\ref{eq_m_multi})). Even though in a bivariate input space the computation demand increases significantly with $D$, the HSGP model works significantly better that the regular GP for most of the non-linear $2D$ functions (even highly wiggly functions) (see Figures B.3-right and C.3 in the online material). However, HSGPs are slower than exact GPs for $D>3$ with a relatively low number of basis functions ($m \gtrsim 5$), as well as even for $D=3$ with a moderate high number of basis functions ($m \gtrsim 20$) in the approximation (see Figure C.3 in the online material). It can be appreciated in all the cases how choosing the optimal boundary factor in the HSGP approximation allows for less number of basis functions (see Figures A.3, B.3-left and C.2 in the online material) and therefore less computational time, which can makes the difference in multivariate input spaces.

%This fact reveals that choosing optimal values for the number of basis functions and boundary factor, looking at the recommendations and diagnosis provided by Figure \ref{fig6_lscale_vs_J_vs_c}, is essential to avoid a excessive computational time, especially in multivariate input spaces.

Roughly similar or even worse consideration is found for splines where serious difficulties have been encountered in building the spline model from $D=3$ and \textit{knots} $> 10$, or even for $D=2$ and \textit{knots} $> 40$ (see Figures B.3-right and C.3 in the online material).}

@@

\noindent The mean function $\mu(t)$ will be defined as an additive model in the form: 
%
\begin{equation} \label{ch5_eq_mean_brithday}
\mu(t) = f_1(t) + f_2(t) + f_3(t) + f_4(t).
\end{equation}

The component $f_1(t)$ represents the long-term trends modeled by a GP with squared exponential covariance function,
%
\begin{equation*}
f_1(t) \sim \mathcal{GP}(0,k_1), \hspace{5mm} k_1(t,t') = \sigma_1^2 \exp\!\!\Big(\!-\frac{1}{2} \frac{(t-t')^2}{\ell_1^2}\Big), 
\end{equation*}

@@

The component $f_4(t)$ represents the special days effects, modeled as a horse-shoe prior model \citep{piironen2017sparsity}:
%
\begin{equation*}
f_4(t) \sim \mathcal{N}(0,\lambda^2_t \tau^2), \hspace{5mm} \lambda^2_t \sim \mathcal{C}^{+}(0,1).
\end{equation*}
